---
title: "Assessing selection bias in the validation dataset"
format: html
toc: true
editor: visual
code-fold: true
warning: false 
message: false
self-contained: true
tabset: true
---

## Introduction

### Note from reviewers

\*Selection bias in the validation dataset (Section 2.3, lines 233-257; Section 5, lines 531-552).

The well-testing dataset is based on voluntary and opportunistic sample submissions. The authors rightly acknowledge this in the Discussion, but the implications for the validity of the entire supervised component of the framework are insufficiently addressed. The supervised step (LASSO feature selection and logistic regression) is trained on these same biased labels. If the probability of testing is correlated with any of the predictor variables (e.g., areas with higher social capacity may test more), this creates a fundamental endogeneity problem that cannot be resolved simply by noting the limitation. The authors should:

-   Provide a descriptive comparison of the spatial and sociodemographic characteristics of tested versus untested grid cells to assess the degree of selection.

-   Discuss how training on biased labels may have influenced which PCs were selected and the resulting index weights.

-   Consider a sensitivity analysis (e.g., varying the assumed contamination rate among untested wells).

## Preprocessing
```{r}
# Load packages
library(sf)
library(dplyr)
library(ggplot2)
library(purrr)
library(stringr)

# Load data
test_sites <- st_read('well_tests_filtered.gpkg')
index_c <- st_read('nc_grid_capacity_imputed.gpkg')
index_h <- st_read('nc_grid_hazard_imputed.gpkg')
index_v <- st_read('nc_grid_vulnerability_imputed.gpkg')

# Align CRS
index_c <- st_transform(index_c, st_crs(test_sites))
index_h <- st_transform(index_h, st_crs(test_sites))
index_v <- st_transform(index_v, st_crs(test_sites))
```

### Extracting boundaries of the target counties
```{r}
target_counties <- c("Alexander County", "Alleghany County","Ashe County","Avery County",
                      "Buncombe County","Burke County","Caldwell County","Catawba County",
                      "Clay County","Cleveland County","Gaston County","Haywood County",
                      "Henderson County","Jackson County","Lincoln County","Macon County",
                      "Madison County","Mcdowell County","Mitchell County","Polk County",
                      "Rutherford County","Transylvania County","Watauga County","Wilkes County",
                      "Yancey County")

NC_counties <- tigris::counties(state = "NC", year = 2024)
western_counties <- NC_counties %>% filter(NAMELSAD %in% target_counties)

index_c <- index_c %>% mutate(county_raw = str_extract(block_group_name, "[A-Za-z .’-]+ County"),
                          county_norm = str_squish(str_to_title(county_raw)))
index_capacity <- index_c %>% filter(!is.na(county_norm) & county_norm %in% target_counties)

#index_h$block_group_name <- index_c$block_group_name
#index_h <- index_h %>% mutate(county_raw = str_extract(block_group_name, "[A-Za-z .’-]+ County"),
#                          county_norm = str_squish(str_to_title(county_raw)))
#index_hazard <- index_h %>% filter(!is.na(county_norm) & county_norm %in% target_counties)

western_counties <- st_transform(western_counties, st_crs(test_sites))

index_hazard <- st_filter(index_h, western_counties, .predicate = st_within)

#index_v <- index_v %>% mutate(county_raw = str_extract(block_group_name, "[A-Za-z .’-]+ County"),
#                          county_norm = str_squish(str_to_title(county_raw)))
#index_capacity <- index_v %>% filter(!is.na(county_norm) & county_norm %in% target_counties)

western_counties <- st_transform(western_counties, st_crs(test_sites))

index_vulnerability <- st_filter(index_v, western_counties, .predicate = st_within)
```

```{r}
# Join index gpkgs and test_sites by geometry
index_capacity <- st_join(index_capacity, test_sites[, c("EIN")], join = st_intersects)
index_capacity$tested <- !is.na(index_capacity$EIN)

index_hazard <- st_join(index_hazard, test_sites[, c("EIN")], join = st_intersects)
index_hazard$tested <- !is.na(index_hazard$EIN)

index_vulnerability <- st_join(index_vulnerability, test_sites[, c("EIN")], join = st_intersects)
index_vulnerability$tested <- !is.na(index_vulnerability$EIN)
```

## Statistical tests

Wilcoxon rank sum test for continuous variables and chi-square test for categorical variables

### Capacity module

```{r}
results_capacity <- list()

# The full range of variables is 4:26
# The table below excludes pct_women, pct_single_parent, pct_under_17, pct_woman_gave_birth, pct_us_born, pct_active_commuting, pct_no_move, pct_crowded_housing, pct_no_computer, housing_units_density
for(i in c(4, 6, 9, 10, 11, 13, 14, 16, 18, 21, 22, 23, 24)) {
  col_name <- names(index_capacity)[i]
  results_capacity[[col_name]] <- wilcox.test(index_capacity[[col_name]] ~ index_capacity$tested)
}

summary_capacity <- imap_dfr(results_capacity, function(res, name) {
  data.frame(
    Variable = name,
    Mean = mean(index_capacity[[name]]),
    SD = sd(index_capacity[[name]]),
    Q1 = quantile(index_capacity[[name]], 0.25),
    Q3 = quantile(index_capacity[[name]], 0.75),
    W_stat = res$statistic,
    p_value = res$p.value
  )
})

# p-value adjustment with Benjamini-Hochberg
p_values <- sapply(results_capacity, function(col) col$p.value)
adj_p <- p.adjust(p_values, method = "BH")

summary_capacity$Adj_p_value = adj_p
summary_capacity$Significant = ifelse(summary_capacity$Adj_p_value < 0.05, "*", "")
rownames(summary_capacity) <- NULL

print(summary_capacity)
```

### Hazard module

```{r}
# Continuous variables
results_hazard_cont <- list()

for(i in 3:9) {
  col_name <- names(index_hazard)[i]
  results_hazard_cont[[col_name]] <- wilcox.test(index_hazard[[col_name]] ~ index_hazard$tested)
}

summary_hazard_cont <- imap_dfr(results_hazard_cont, function(res, name) {
  data.frame(
    Variable = name,
    Mean = mean(index_hazard[[name]]),
    SD = sd(index_hazard[[name]]),
    Q1 = quantile(index_hazard[[name]], 0.25),
    Q3 = quantile(index_hazard[[name]], 0.75),
    W_stat = res$statistic,
    p_value = res$p.value
  )
})

# p-value adjustment with Benjamini-Hochberg
p_values <- sapply(results_hazard_cont, function(col) col$p.value)
adj_p <- p.adjust(p_values, method = "BH")

summary_hazard_cont$Adj_p_value = adj_p
summary_hazard_cont$Significant = ifelse(summary_hazard_cont$Adj_p_value < 0.05, "*", "")
rownames(summary_hazard_cont) <- NULL

print(summary_hazard_cont)

# Categorical variables
results_hazard_cat <- list()

for(i in c(10)) {
  index_hazard[[col_name]] <- as.character(index_hazard[[col_name]])
  col_name <- names(index_hazard)[i]
  results_hazard_cat[[col_name]] <- chisq.test(table(index_hazard$tested, index_hazard[[col_name]]))
}

get_mode <- function(x) {
  unique_x <- unique(x)
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

summary_hazard_cat <- imap_dfr(results_hazard_cat, function(res, name) {
  data.frame(
    Variable = name,
    Mode = get_mode(index_hazard[[name]]),
    chi_square = res$statistic,
    p_value = res$p.value
  )
})

# p-value adjustment with Benjamini-Hochberg
p_values <- sapply(results_hazard_cat, function(col) col$p.value)
adj_p <- p.adjust(p_values, method = "BH")

summary_hazard_cat$Adj_p_value = adj_p
summary_hazard_cat$Significant = ifelse(summary_hazard_cat$Adj_p_value < 0.05, "*", "")
rownames(summary_hazard_cat) <- NULL

print(summary_hazard_cat)
```

### Vulnerability module

```{r}
# Variable indices
cont_indices <- c(6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41, 42, 43, 47)

cat_indices <- c(3, 4, 5, 39, 45, 48)

# Continuous variables
results_vulnerability_cont <- list()

for(i in cont_indices) {
  col_name <- names(index_vulnerability)[i]
  results_vulnerability_cont[[col_name]] <- wilcox.test(index_vulnerability[[col_name]] ~ index_vulnerability$tested)
}

summary_vulnerability_cont <- imap_dfr(results_vulnerability_cont, function(res, name) {
  data.frame(
    Variable = name,
    Mean = mean(index_vulnerability[[name]]),
    SD = sd(index_vulnerability[[name]]),
    Q1 = quantile(index_vulnerability[[name]], 0.25),
    Q3 = quantile(index_vulnerability[[name]], 0.75),
    W_stat = res$statistic,
    p_value = res$p.value
  )
})

# p-value adjustment with Benjamini-Hochberg
p_values <- sapply(results_vulnerability_cont, function(col) col$p.value)
adj_p <- p.adjust(p_values, method = "BH")

summary_vulnerability_cont$Adj_p_value = adj_p
summary_vulnerability_cont$Significant = ifelse(summary_vulnerability_cont$Adj_p_value < 0.05, "*", "")
rownames(summary_vulnerability_cont) <- NULL

print(summary_vulnerability_cont)

# Categorical variables
results_vulnerability_cat <- list()

for(i in cat_indices) {
  index_vulnerability[[col_name]] <- as.character(index_vulnerability[[col_name]])
  col_name <- names(index_vulnerability)[i]
  results_vulnerability_cat[[col_name]] <- chisq.test(table(index_vulnerability$tested, index_vulnerability[[col_name]]))
}

summary_vulnerability_cat <- imap_dfr(results_vulnerability_cat, function(res, name) {
  data.frame(
    Variable = name,
    Mode = get_mode(index_vulnerability[[name]]),
    chi_square = res$statistic,
    p_value = res$p.value
  )
})

# p-value adjustment with Benjamini-Hochberg
p_values <- sapply(results_vulnerability_cat, function(col) col$p.value)
adj_p <- p.adjust(p_values, method = "BH")

summary_vulnerability_cat$Adj_p_value = adj_p
summary_vulnerability_cat$Significant = ifelse(summary_vulnerability_cat$Adj_p_value < 0.05, "*", "")
rownames(summary_vulnerability_cat) <- NULL

print(summary_vulnerability_cat)
```

### All modules
```{r}
summary_all_cont <- rbind(summary_capacity, summary_hazard_cont, summary_vulnerability_cont)
print(summary_all_cont)

summary_all_cat <- rbind(summary_hazard_cat, summary_vulnerability_cat)
print(summary_all_cat)
```