---
title: "Validation of overall risk index using NCDHHS well testing results for E. coli and Total Coliform"
format: html
toc: true
editor: visual
code-fold: true
warning: false 
message: false
self-contained: true
tabset: true
---

## Introduction

Recall that we applied constrained PCA to obtain principal components (PCs) whose directions align with the expected relationship to contamination risk. Constraints were only imposed on variables with known directional associations (positive or negative) with risk; variables without such expectations were left unconstrained. Note that we did not apply varimax rotation as such rotation does not preserve the sign constraints we impose.

In this report, we use the validation dataset to guide the construction of a composite risk score based on these PCs. The analysis proceeds as follows. We first divide the validation dataset into training and testing subsets. The outcome variable is binary, indicating total coliform detection (detect vs. non-detect). Using the training data, we conduct two analyses.

1.  **Model comparison**: We fit three random-forest models: (a) a model that uses only the PCs, (b) a model that uses all original variables from the physical vulnerability, social capacity, and hazard modules, and (c) a model that uses only the first PC from each module. If models (a) and (b) exhibit similar performance, this provides evidence that constrained PCA effectively preserves the relevant predictive information. If both (a) and (b) outperform model (c), this indicates that relying solely on the first PC from each module is insufficient for explaining contamination risk.

2.  **Index construction**: To derive a one-dimensional index for each module, we apply the LASSO method to determine the optimal weights for the PCs based on contamination risk. The resulting weighted combination of PCs is standardized to form a numerical index for each module.

Finally, we compute the composite contamination risk score as $$
\text{Risk} = \text{Physical Vulnerability} + \text{Hazard} - \text{Social Capacity}
$$ We then use the testing dataset to conduct a two-sample test comparing the mean (or central tendency) of the risk score between detect and non-detect groups.

## Data Preprocessing

We perform a stratified sampling procedure to split the validation dataset into training and testing subsets in a 1:1 ratio, ensuring that both subsets preserve the same proportion of positive and negative coliform samples as in the original dataset.

```{r}
# Load packages
library(sf)
library(dplyr)
library(ggplot2)
library(lubridate)
library(changepoint)
library(ROCit)
library(cutpointr)
library(rpart)
library(rpart.plot)
library(arm)
library(caret)
library(glmnet)
library(pROC)
# read data
data <- st_read("validation.gpkg")
data$time <- as.numeric(data$Date_Collected - as.Date("2024-09-25"))
vul_dat <- st_read("index_vulnerability_detailed.gpkg")
haz_dat <- st_read("index_hazard_detailed.gpkg")
cap_dat <- st_read("index_capacity_detailed.gpkg")

# ---------------------------
# Outcome definition
# ---------------------------
data$coliform <- ifelse(data$Coliform_Result_Binary == "Present", 1, 0)

vul_tab <- sf::st_drop_geometry(vul_dat)
haz_tab <- sf::st_drop_geometry(haz_dat)
cap_tab <- sf::st_drop_geometry(cap_dat)

merged_data <- data %>%
  dplyr::left_join(vul_tab, by = "grid_id") %>%
  dplyr::left_join(haz_tab, by = "grid_id") %>%
  dplyr::left_join(cap_tab, by = "grid_id") %>%
  # remove duplicate join suffixes
  dplyr::select(
    -tidyselect::matches("\\.x$"),
    -tidyselect::matches("\\.y$"),
    -tidyselect::matches("\\.x\\.x$"),
    -tidyselect::matches("\\.y\\.y$")
  ) %>%
  # drop obvious non-predictors / outcomes
  dplyr::select(
    -tidyselect::any_of(c(
      "System_Name", "Street_Address", "City", "State", "Zip",
      "Street_Address_Concat", "EIN", "StarLiMS_ID",
      "Date_Collected", "Sampling_Point", "Comment",
      "Accuracy", "Accuracy_Type", "grid_id",
      "index_vulnerability", "index_hazard", "index_capacity",
      "risk", "Total_Coliform_Result", "Ecoli_Result"
    ))
  ) %>%
  # drop geometry if still present
  sf::st_drop_geometry() %>%
  # drop all A_* variables and specific unwanted columns
  dplyr::select(
    -tidyselect::starts_with("A_"),
    -tidyselect::any_of(c(
      "Ecoli_Result_Binary",
      "Coliform_Result_Binary",
      "ecoli",
      "y_fake"
    ))
  )

# ---------------------------
# Remove all-NA / constant columns (helps fitting)
# ---------------------------
is_all_na   <- vapply(merged_data, function(x) all(is.na(x)), logical(1))
is_const    <- vapply(merged_data, function(x) length(unique(na.omit(x))) <= 1, logical(1))
merged_data <- merged_data[, !(is_all_na | is_const), drop = FALSE]

# Characters -> factors; ensure binary 0/1 outcome
merged_data[] <- lapply(merged_data, function(x) if (is.character(x)) factor(x) else x)
stopifnot("coliform" %in% names(merged_data))
merged_data$coliform <- as.integer(merged_data$coliform == 1)
stopifnot(length(unique(merged_data$coliform)) == 2)

# ================================================================
# 4:1 STRATIFIED TRAIN-TEST SPLIT (80/20)
# ================================================================
# set.seed(218)
set.seed(403)
train_idx <- caret::createDataPartition(merged_data$coliform, p = 0.6, list = FALSE)
train_dat <- merged_data[train_idx, , drop = FALSE]
test_dat  <- merged_data[-train_idx, , drop = FALSE]

cat(sprintf("Training size: %d | Testing size: %d\n", nrow(train_dat), nrow(test_dat)))
cat(sprintf("Class balance (train): %.2f%% positives\n", 100 * mean(train_dat$coliform)))
cat(sprintf("Class balance (test):  %.2f%% positives\n", 100 * mean(test_dat$coliform)))
```

## Model Comparison

Next, we compare three random-forest models. They differ in the covariates used.

1.  The first model uses the principal components (PCs) from each module, where the number of PCs is selected to explain at least 80% of the variance within that module. This results in 29 PCs for physical vulnerability, 10 PCs for social capacity, and 6 PCs for hazard.

2.  The second model uses all original variables from the three modules without applying dimensionality reduction.

3.  The third model uses only the first PC from each module. We refer to this as the “unsupervised’’ approach because these individual PCs are used directly as composite indices, similar to the construction of a social vulnerability index, and their formation is not guided by supervised learning.

Each model also includes the number of days since Hurricane Helene affected North Carolina.

For each model, we first train an initial random forest using default settings to obtain variable importance scores. To reduce noise and focus on the most informative predictors, we retain only variables whose importance is at or above the 25th percentile of the importance distribution. A refined random forest is then trained using only these screened predictors. Hyperparameters---specifically, the number of variables randomly sampled at each split (\\texttt{mtry}) and the minimum terminal node size (\\texttt{nodesize})---are tuned using 5-fold cross-validation, with the optimal combination selected based on the highest cross-validated area under the ROC curve (AUC). A final tuned random forest is then fitted on the full training dataset, and the decision threshold is chosen to maximize the F1-score on the training data.

We compare the three models using the AUC computed on the testing dataset. If the PC-based model achieves an AUC comparable to that of the model using all original variables, this indicates that the dimensionality reduction via PCA preserves most of the predictive information relevant for modeling contamination risk.

The results for the area under the curve (AUC) and the receiver operating characteristic (ROC) curve are summarized below. The model using only the principal components (PCs) achieves an AUC of 0.645 and an F1-score of 0.449, while the model using all original variables attains an AUC of 0.621 and an F1-score of 0.421. These results suggest that the dimensionality-reduced representation retains most of the predictive information.\

In comparison, the unsupervised approach yields an AUC of 0.590 and an F1-score of 0.415. Relative to both the PC-only and non-PC models, this indicates that relying solely on the first PC from each module is not sufficient for predicting contamination risk.

```{r}
###############################################################################
# Tuned Random Forest (randomForest only)
#
# - PC-only vs Non-PC vs Unsupervised (haz_PC1, vul_PC1, cap_PC1, time)
# - Step 1: Align TRAIN/TEST, impute, and harmonize factor levels
# - Step 2: Stratified k-fold CV to tune (mtry, nodesize) by AUC
# - Step 3: Fit final RF with tuned hyperparameters
# - Step 4: Choose probability cutoff via F1 on TRAIN
# - Randomness in CV folds and RF training is controlled by `rf_seed`
# - Report Train/Test AUC & F1; plot TEST ROC curves
###############################################################################

suppressPackageStartupMessages({
  library(randomForest)
  library(pROC)
})

# ================================================================
# Helpers: alignment / preprocessing
# ================================================================
coerce_numeric_like <- function(x) {
  if (is.factor(x)) {
    suppressWarnings({
      y <- as.numeric(as.character(x))
    })
    if (!all(is.na(y))) return(y)
  }
  x
}

fct_mode <- function(f) {
  tab <- table(f)
  names(tab)[which.max(tab)]
}

normalize_empty_to_na <- function(x) {
  if (!is.character(x)) return(x)
  x_trim <- trimws(x)
  x_trim[x_trim %in% c("", "NA", "N/A", "na", "n/a", "<NA>")] <- NA_character_
  x_trim
}

# Align TRAIN and TEST on a given set of predictors, with "__OTHER__" bucket for new test levels
impute_and_align <- function(train, test, vars,
                             outcome = "coliform",
                             other   = "__OTHER__") {
  # Only keep predictors that exist in BOTH sets
  use_vars <- intersect(vars, intersect(names(train), names(test)))
  if (length(use_vars) == 0L)
    stop("No overlapping variables between train/test for alignment.")

  tr <- train[, c(outcome, use_vars), drop = FALSE]
  te <- test[,  c(outcome, use_vars), drop = FALSE]

  # Coerce numeric-like factors to numeric
  tr[] <- lapply(tr, coerce_numeric_like)
  te[] <- lapply(te, coerce_numeric_like)

  # Normalize empty strings in character columns
  tr[] <- lapply(tr, function(x) if (is.character(x)) normalize_empty_to_na(x) else x)
  te[] <- lapply(te, function(x) if (is.character(x)) normalize_empty_to_na(x) else x)

  common <- setdiff(intersect(names(tr), names(te)), outcome)

  for (v in common) {
    if (is.numeric(tr[[v]]) || is.integer(tr[[v]])) {
      # numeric: median imputation
      med <- suppressWarnings(median(tr[[v]], na.rm = TRUE))
      if (!is.finite(med)) med <- 0
      tr[[v]][is.na(tr[[v]])] <- med
      te[[v]][is.na(te[[v]])] <- med

      if (!(is.numeric(te[[v]]) || is.integer(te[[v]]))) {
        suppressWarnings(te[[v]] <- as.numeric(as.character(te[[v]])))
        te[[v]][is.na(te[[v]])] <- med
      }

    } else {
      # treat as categorical
      tr_chr <- as.character(tr[[v]])
      te_chr <- as.character(te[[v]])

      tr_mode <- fct_mode(factor(tr_chr))
      tr_chr[is.na(tr_chr)] <- tr_mode
      te_chr[is.na(te_chr)] <- tr_mode

      L_train <- sort(unique(tr_chr))
      # map unseen test levels to OTHER
      te_chr[!(te_chr %in% L_train)] <- other
      L_full <- unique(c(L_train, other))

      tr[[v]] <- factor(tr_chr, levels = L_full)
      te[[v]] <- factor(te_chr, levels = L_full)
    }
  }

  # Drop zero-variance columns in TRAIN (and mirror in TEST)
  nzv <- vapply(tr[, common, drop = FALSE], function(x) {
    if (is.numeric(x) || is.integer(x)) sd(x, na.rm = TRUE) > 0
    else if (is.factor(x)) length(unique(x)) > 1
    else TRUE
  }, logical(1))

  keep <- names(which(nzv))
  tr <- tr[, c(outcome, keep), drop = FALSE]
  te <- te[, c(outcome, keep), drop = FALSE]

  list(train = tr, test = te)
}

# ================================================================
# Outcome & metric helpers
# ================================================================
# Encode outcome as "pos"/"neg"
make_posneg <- function(y) {
  y_chr   <- as.character(y)
  pos_vals <- c("1","TRUE","T","Yes","Y","pos","Pos","positive","case")
  factor(ifelse(y_chr %in% pos_vals, "pos", "neg"), levels = c("pos","neg"))
}

# For pROC: negative first, positive second
make_negpos_for_pROC <- function(y_posneg) {
  factor(ifelse(y_posneg == "pos", "pos", "neg"), levels = c("neg","pos"))
}

roc_auc_pair <- function(y_negpos, prob_pos) {
  r <- pROC::roc(response = y_negpos, predictor = prob_pos,
                 levels = c("neg","pos"), quiet = TRUE)
  list(roc = r, auc = as.numeric(pROC::auc(r)))
}

f1_score <- function(y_true, y_pred) {
  # y_true, y_pred: "pos"/"neg"
  tp <- sum(y_true == "pos" & y_pred == "pos")
  fp <- sum(y_true == "neg" & y_pred == "pos")
  fn <- sum(y_true == "pos" & y_pred == "neg")

  precision <- ifelse(tp + fp == 0, 0, tp / (tp + fp))
  recall    <- ifelse(tp + fn == 0, 0, tp / (tp + fn))

  if (precision + recall == 0) return(0)
  2 * precision * recall / (precision + recall)
}

# Choose the probability cutoff that maximizes F1 on TRAIN
optimal_f1_cutoff <- function(y_true_posneg, prob_pos) {
  y_true_posneg <- factor(y_true_posneg, levels = c("pos","neg"))
  cuts    <- seq(0.01, 0.99, by = 0.01)
  f1_vals <- numeric(length(cuts))

  for (i in seq_along(cuts)) {
    pred <- ifelse(prob_pos >= cuts[i], "pos", "neg")
    f1_vals[i] <- f1_score(y_true_posneg, pred)
  }

  cuts[which.max(f1_vals)]
}

# ================================================================
# Stratified k-folds (manual, no caret)
# ================================================================
stratified_folds <- function(y, k = 5, seed = 123) {
  set.seed(seed)
  y  <- factor(y)
  lv <- levels(y)
  folds <- vector("list", k)
  for (lev in lv) {
    idx <- which(y == lev)
    idx <- sample(idx, length(idx))
    parts <- split(idx, rep(1:k, length.out = length(idx)))
    for (i in seq_len(k)) {
      folds[[i]] <- c(folds[[i]], parts[[i]])
    }
  }
  lapply(folds, sort)
}

# ================================================================
# CV tuning for randomForest: tune mtry & nodesize by AUC
# Randomness controlled by `rf_seed`
# ================================================================
cv_tune_rf <- function(train_df, mtry_grid, nodesize_grid,
                       k = 5, ntree = 2000,
                       quiet = FALSE, rf_seed = 123) {

  y     <- make_posneg(train_df$coliform)
  folds <- stratified_folds(y, k = k, seed = rf_seed)

  p <- ncol(train_df) - 1L
  mtry_grid     <- sort(unique(pmax(1L, pmin(p, as.integer(mtry_grid)))))
  nodesize_grid <- sort(unique(as.integer(nodesize_grid)))

  res_tbl <- data.frame(
    mtry     = integer(),
    nodesize = integer(),
    mean_auc = numeric()
  )

  for (mtry_g in mtry_grid) {
    for (nodesize_g in nodesize_grid) {

      aucs <- numeric(length(folds))

      for (i in seq_along(folds)) {

        val_idx <- folds[[i]]
        tr_idx  <- setdiff(seq_len(nrow(train_df)), val_idx)

        tr <- train_df[tr_idx, , drop = FALSE]
        va <- train_df[val_idx, , drop = FALSE]

        tr$coliform <- make_posneg(tr$coliform)
        va$coliform <- make_posneg(va$coliform)

        # RF randomness determined by rf_seed
        set.seed(rf_seed + mtry_g + nodesize_g + i)

        rf <- randomForest(
          coliform ~ ., data = tr,
          ntree    = ntree,
          mtry     = mtry_g,
          nodesize = nodesize_g
        )

        prob_va <- as.numeric(predict(rf, va, type = "prob")[, "pos"])
        roc_res <- roc_auc_pair(make_negpos_for_pROC(va$coliform), prob_va)
        aucs[i] <- roc_res$auc
      }

      mean_auc <- mean(aucs, na.rm = TRUE)

      if (!quiet) {
        cat(sprintf("[CV] seed=%d | mtry=%d nodesize=%d -> AUC=%.3f\n",
                    rf_seed, mtry_g, nodesize_g, mean_auc))
      }

      res_tbl <- rbind(
        res_tbl,
        data.frame(mtry = mtry_g,
                   nodesize = nodesize_g,
                   mean_auc = mean_auc)
      )
    }
  }

  best <- res_tbl[which.max(res_tbl$mean_auc), , drop = FALSE]
  best
}

# ================================================================
# Fit tuned RF using best (mtry, nodesize); cutoff via TRAIN F1
# Randomness controlled by `rf_seed`
# ================================================================
fit_rf_tuned <- function(train_df, test_df,
                         best_mtry, best_nodesize,
                         ntree = 2000, quiet = FALSE,
                         rf_seed = 123) {

  train_df$coliform <- make_posneg(train_df$coliform)
  test_df$coliform  <- make_posneg(test_df$coliform)

  # RF seed controls final model randomness
  set.seed(rf_seed)

  rf_mod <- randomForest(
    coliform ~ ., data = train_df,
    ntree    = ntree,
    mtry     = best_mtry,
    nodesize = best_nodesize
  )

  prob_tr <- as.numeric(predict(rf_mod, train_df, type = "prob")[, "pos"])
  prob_te <- as.numeric(predict(rf_mod, test_df,  type = "prob")[, "pos"])

  # Cutoff chosen to maximize F1 on TRAIN
  best_cut <- optimal_f1_cutoff(train_df$coliform, prob_tr)

  # AUCs
  roc_tr <- roc_auc_pair(make_negpos_for_pROC(train_df$coliform), prob_tr)
  roc_te <- roc_auc_pair(make_negpos_for_pROC(test_df$coliform),  prob_te)

  # F1 using training-optimal cutoff
  pred_tr <- ifelse(prob_tr >= best_cut, "pos", "neg")
  pred_te <- ifelse(prob_te >= best_cut, "pos", "neg")
  f1_tr   <- f1_score(train_df$coliform, pred_tr)
  f1_te   <- f1_score(test_df$coliform,  pred_te)

  if (!quiet) {
    cat(sprintf(
      "[FINAL RF] seed=%d | mtry=%d nodesize=%d cutoff=%.2f | Train AUC=%.3f, F1=%.3f | Test AUC=%.3f, F1=%.3f\n",
      rf_seed, best_mtry, best_nodesize, best_cut,
      roc_tr$auc, f1_tr,
      roc_te$auc, f1_te
    ))
  }

  list(
    model      = rf_mod,
    auc_train  = roc_tr$auc,
    auc_test   = roc_te$auc,
    f1_train   = f1_tr,
    f1_test    = f1_te,
    prob_train = prob_tr,
    prob_test  = prob_te,
    cutoff     = best_cut,
    roc_train  = roc_tr$roc,
    roc_test   = roc_te$roc
  )
}

# ================================================================
# Helper: extract importance vector (MeanDecreaseGini)
# (currently unused, kept for possible future diagnostics)
# ================================================================
get_importance_vector <- function(rf_obj) {
  imp <- randomForest::importance(rf_obj, type = 2)  # type=2: Gini
  if (is.matrix(imp)) {
    if ("MeanDecreaseGini" %in% colnames(imp)) {
      imp_vec <- imp[, "MeanDecreaseGini"]
    } else {
      imp_vec <- imp[, 1L]
    }
  } else {
    imp_vec <- imp
  }
  imp_vec <- as.numeric(imp_vec)
  names(imp_vec) <- rownames(imp)
  imp_vec
}

# ================================================================
# Run a block: PC-only, Non-PC, or Unsupervised
#   Step 1: Align TRAIN/TEST
#   Step 2: CV tune RF on all aligned predictors
#   Step 3: Fit final tuned RF
#   Randomness controlled by `rf_seed`
# ================================================================
run_block <- function(vars, tag,
                      ntree = 2000,
                      quiet = FALSE,
                      rf_seed = 123) {

  pair <- impute_and_align(train_dat, test_dat, vars)

  train_red <- pair$train
  test_red  <- pair$test

  p <- ncol(train_red) - 1L
  mtry_grid <- unique(round(c(sqrt(p)/2, sqrt(p), 2 * sqrt(p), p / 3)))
  mtry_grid <- pmax(1L, pmin(p, mtry_grid))
  nodesize_grid <- c(1, 3, 5, 10, 20)

  if (!quiet) {
    cat(sprintf("\n===== %s — CV tuning (seed=%d) =====\n", tag, rf_seed))
  }
  best <- cv_tune_rf(train_red, mtry_grid, nodesize_grid,
                     k = 5, ntree = ntree,
                     quiet = quiet, rf_seed = rf_seed)

  if (!quiet) {
    cat(sprintf("Best params for %s: mtry=%d, nodesize=%d (mean CV AUC=%.3f)\n",
                tag, best$mtry, best$nodesize, best$mean_auc))
    cat(sprintf("===== %s — Fit final tuned RF (seed=%d) =====\n", tag, rf_seed))
  }

  fit <- fit_rf_tuned(train_red, test_red,
                      best_mtry     = best$mtry,
                      best_nodesize = best$nodesize,
                      ntree  = ntree,
                      quiet  = quiet,
                      rf_seed = rf_seed)

  list(best = best, fit = fit)
}

# ================================================================
# Feature sets
#   (assumes train_dat/test_dat already defined by a fixed split)
# ================================================================
all_vars   <- setdiff(names(train_dat), "coliform")
pc_vars    <- grep("_PC", all_vars, value = TRUE)
nonpc_vars <- setdiff(all_vars, pc_vars)

# ---- Drop specific social-capacity variables from Non-PC set ----
drop_cap_vars <- c(
  "pct_women",
  "pct_single_parent",
  "pct_under_17",
  "pct_women_gave_birth",
  "pct_us_born",
  "pct_active_commuting",
  "pct_no_move",
  "pct_crowded_housing",
  "pct_no_computer",
  "housing_units_density"
)
nonpc_vars <- setdiff(nonpc_vars, drop_cap_vars)

# ---- PC-only: time + all PCs ----
pc_vars <- c("time", pc_vars)

# ---- Unsupervised: haz_PC1, vul_PC1, cap_PC1, time ----
unsup_vars <- c("time", "haz_PC1", "vul_PC1", "cap_PC1")

# ================================================================
# Run tuned RF for PC-only, Non-PC, and Unsupervised with chosen RF seed
#   (RF randomness only; TRAIN/TEST split assumed fixed elsewhere)
# ================================================================
rf_seed_use <- 166

res_pc <- run_block(pc_vars,    tag = "PC-only",
                    ntree = 2000, quiet = TRUE,
                    rf_seed = rf_seed_use)

res_nonpc <- run_block(nonpc_vars, tag = "Non-PC",
                       ntree = 2000, quiet = TRUE,
                       rf_seed = rf_seed_use)

res_unsup <- run_block(unsup_vars, tag = "Unsupervised",
                       ntree = 2000, quiet = TRUE,
                       rf_seed = rf_seed_use)

# ================================================================
# Final TEST ROC comparison: PC-only vs Non-PC vs Unsupervised
# ================================================================
if (!is.null(res_pc) && !is.null(res_nonpc) && !is.null(res_unsup)) {
  plot(res_pc$fit$roc_test,
       col = "magenta", lwd = 2,
       main = "TEST ROC: PC-only vs Non-PC vs Unsupervised",
       legacy.axes = TRUE)

  plot(res_nonpc$fit$roc_test,
       col = "deepskyblue", lwd = 2, add = TRUE)

  # Unsupervised with indianred1
  plot(res_unsup$fit$roc_test,
       col = "indianred1", lwd = 2, add = TRUE)

  abline(a = 0, b = 1, lty = 2)

  legend("bottomright",
         legend = c(
           sprintf("PC-only        (AUC=%.3f, F1=%.3f, c=%.2f)",
                   res_pc$fit$auc_test,    res_pc$fit$f1_test,    res_pc$fit$cutoff),
           sprintf("Non-PC         (AUC=%.3f, F1=%.3f, c=%.2f)",
                   res_nonpc$fit$auc_test, res_nonpc$fit$f1_test, res_nonpc$fit$cutoff),
           sprintf("Unsupervised   (AUC=%.3f, F1=%.3f, c=%.2f)",
                   res_unsup$fit$auc_test, res_unsup$fit$f1_test, res_unsup$fit$cutoff)
         ),
         col = c("magenta", "deepskyblue", "indianred1"),
         lwd = 2, bty = "n")
}
```

## Index Construction

Next, we apply logistic regression with a LASSO penalty to construct composite risk scores for the different modules. This framework is implemented in two ways: (1) using the PCs as predictors and (2) using the original non-PC variables. The tuning parameter $\lambda$ is selected via cross-validation based on the area under the ROC curve (AUC). Specifically, we choose the largest $\lambda$ whose cross-validated AUC lies within one standard error of the maximum (the \`\`1-SE rule''), resulting in a sparser yet near-optimal model.

### Supervised (PCs)

We apply a logistic regression model with a LASSO penalty to estimate the relative weights of the PCs within each module. Sign constraints are imposed on the coefficients to ensure interpretability (positive for vulnerability and hazard, negative for capacity). To further enhance interpretability, we then choose the value of $\lambda$ along this path whose model size, measured by the number of nonzero coefficients, is closest to a target of five active PCs per module.

To reduce sensitivity to any single train–test split, we repeat the logistic LASSO procedure 500 times across independent random splits within each module and retain only those PCs that are selected in more than half of the iterations through majority voting. Using the selected PCs within each module, we construct module-specific indices by forming weighted sums of these PCs, where the weights are obtained from sign-constrained logistic regression fitted on the training data.

```{r}
## ================================================================
## PACKAGES
## ================================================================
suppressPackageStartupMessages({
  if (!requireNamespace("glmnet", quietly = TRUE))
    install.packages("glmnet", repos = "https://cloud.r-project.org")
  if (!requireNamespace("pROC", quietly = TRUE))
    install.packages("pROC", repos = "https://cloud.r-project.org")
  if (!requireNamespace("caret", quietly = TRUE))
    install.packages("caret", repos = "https://cloud.r-project.org")
  if (!requireNamespace("dplyr", quietly = TRUE))
    install.packages("dplyr", repos = "https://cloud.r-project.org")
  library(glmnet)
  library(pROC)
  library(caret)
  library(dplyr)
})

## ================================================================
## SIMPLE LEAKAGE-SAFE IMPUTATION
## ================================================================
impute_full <- function(df, vars, outcome = "coliform") {
  out <- df[, c(outcome, intersect(vars, names(df))), drop = FALSE]
  out[] <- lapply(out, function(x) if (is.character(x)) factor(x) else x)
  
  for (v in setdiff(names(out), outcome)) {
    if (is.numeric(out[[v]])) {
      med <- suppressWarnings(stats::median(out[[v]], na.rm = TRUE))
      if (!is.finite(med)) med <- 0
      out[[v]][is.na(out[[v]])] <- med
    } else {
      if (!is.factor(out[[v]])) out[[v]] <- factor(out[[v]])
      lev <- levels(out[[v]])
      if (!("Missing" %in% lev)) lev <- c(lev, "Missing")
      out[[v]] <- factor(out[[v]], levels = lev)
      out[[v]][is.na(out[[v]])] <- "Missing"
    }
  }
  out
}

## ================================================================
## HELPER — COUNT NONZERO COEFFS (EXCLUDE INTERCEPT)
## ================================================================
count_nonzero <- function(beta_mat) {
  sum(beta_mat[rownames(beta_mat) != "(Intercept)", 1] != 0)
}

## ================================================================
## CORE FITTER: CONSTRAINED LOGISTIC LASSO ON PCs
## ================================================================
fit_lasso_module_pc_only <- function(
    df, module_vars, outcome = "coliform",
    pc_pattern = "PC", ignore_case = FALSE,
    module_name = "module",
    nfolds = 10,
    lower.limits = -Inf,
    upper.limits =  Inf,
    # sparsity controls
    dfmax = NULL, pmax = NULL,
    penalty.factor = NULL,   # can be named; aligns safely below
    use_1se = TRUE,
    target_k = NULL          # try to pick λ giving ~k nonzeros
) {
  stopifnot(outcome %in% names(df))
  
  # keep only PC variables present
  keep <- intersect(module_vars, names(df))
  keep <- setdiff(keep, outcome)
  keep <- keep[grepl(pc_pattern, keep, ignore.case = ignore_case)]
  if (length(keep) == 0L) {
    message(sprintf("[%s] no PC variables found; skipping.", module_name))
    return(NULL)
  }
  
  dat_mod <- impute_full(df, keep, outcome = outcome)
  
  # Design matrix; drop intercept
  X <- model.matrix(as.formula(paste(outcome, "~ .")), data = dat_mod)[, -1, drop = FALSE]
  y <- dat_mod[[outcome]]
  
  # drop constant columns
  is_const <- vapply(seq_len(ncol(X)), function(j) length(unique(X[, j])) <= 1, logical(1))
  if (any(is_const)) X <- X[, !is_const, drop = FALSE]
  if (ncol(X) == 0L) {
    message(sprintf("[%s] no nonconstant predictors; skipping.", module_name))
    return(NULL)
  }
  
  # ----- Robust alignment of penalty.factor to colnames(X) -----
  if (is.null(penalty.factor)) {
    penalty.factor <- rep(1, ncol(X))
  } else {
    pf <- penalty.factor
    if (!is.null(names(pf))) {
      pf_aligned <- pf[colnames(X)]
      pf_aligned[is.na(pf_aligned)] <- 1
      penalty.factor <- as.numeric(pf_aligned)
    } else if (length(pf) == ncol(X)) {
      penalty.factor <- as.numeric(pf)
    } else {
      pc_id <- suppressWarnings(as.integer(sub(".*PC(\\d+).*", "\\1", colnames(X))))
      if (all(!is.na(pc_id))) {
        rng <- range(pc_id, na.rm = TRUE)
        penalty.factor <- 1 + (pc_id - rng[1]) / (diff(rng) + 1e-8)
      } else {
        penalty.factor <- rep(1, ncol(X))
      }
    }
  }
  
  set.seed(666)
  cvfit <- cv.glmnet(
    x = X, y = y, family = "binomial",
    alpha = 1, standardize = TRUE, nfolds = nfolds,
    type.measure = "auc",
    lower.limits = lower.limits, upper.limits = upper.limits,
    dfmax = dfmax, pmax = pmax,
    penalty.factor = penalty.factor
  )
  
  # Prefer sparser lambda.1se when available
  lam_base <- if (use_1se && !is.null(cvfit$lambda.1se)) cvfit$lambda.1se else cvfit$lambda.min
  
  # Optionally seek λ with ≈ target_k nonzeros
  if (!is.null(target_k)) {
    all_lam <- sort(unique(c(cvfit$lambda, lam_base)), decreasing = TRUE) # larger λ → sparser
    nz_count <- vapply(all_lam, function(s) {
      b <- as.matrix(coef(cvfit$glmnet.fit, s = s))
      count_nonzero(b)
    }, integer(1))
    idx <- which.min(abs(nz_count - target_k))
    lam_base <- all_lam[idx]
  }
  
  # AUCs (TRAIN) for lambda.min and lam_base
  p_min <- as.numeric(predict(cvfit, newx = X, s = "lambda.min", type = "response"))
  roc_m  <- pROC::roc(y, p_min, quiet = TRUE)
  auc_m  <- as.numeric(pROC::auc(roc_m))
  
  p_sel <- as.numeric(predict(cvfit, newx = X, s = lam_base, type = "response"))
  roc_b <- pROC::roc(y, p_sel, quiet = TRUE)
  auc_b <- as.numeric(pROC::auc(roc_b))
  
  # Coefficients at lam_base
  bb <- as.matrix(coef(cvfit, s = lam_base))
  nz <- bb[rownames(bb) != "(Intercept)" & as.vector(bb) != 0, , drop = FALSE]
  nz <- nz[order(abs(nz[, 1]), decreasing = TRUE), , drop = FALSE]
  
  list(
    module     = module_name,
    cvfit      = cvfit,
    X          = X,
    y          = y,
    lambda_sel = lam_base,
    roc_min    = roc_m,
    auc_min    = auc_m,
    roc_best   = roc_b,
    auc_best   = auc_b,
    coef_best  = nz
  )
}

## ================================================================
## EXTEND PATH IF CURRENT λ GIVES ALL-ZERO
## ================================================================
pick_lambda_per_module <- function(
    res,
    lower.limits,
    upper.limits,
    nlambda_extend = 400,
    lambda_min_ratio_extend = 1e-6
) {
  if (is.null(res)) return(NULL)
  if (!is.null(res$coef_best) && nrow(res$coef_best) > 0) return(res)
  
  fit_ext <- glmnet(
    x = res$X, y = res$y,
    family = "binomial", alpha = 1, standardize = TRUE,
    nlambda = nlambda_extend,
    lambda.min.ratio = lambda_min_ratio_extend,
    lower.limits = lower.limits,
    upper.limits = upper.limits
  )
  
  lam_best <- NA_real_
  nz <- NULL
  
  # search for first λ (from small to large) with nonzero coefficients
  for (lam in sort(fit_ext$lambda, decreasing = FALSE)) {
    b <- as.matrix(coef(fit_ext, s = lam))
    b <- b[rownames(b) != "(Intercept)" & as.vector(b) != 0, , drop = FALSE]
    if (nrow(b) > 0) {
      lam_best <- lam
      nz <- b
      break
    }
  }
  
  if (is.na(lam_best)) {
    message(sprintf("[%s] no nonzero PCs even after extension.", res$module))
    res$lambda_sel <- NA_real_
    res$coef_best  <- NULL
    return(res)
  }
  
  p <- as.numeric(predict(fit_ext, newx = res$X, s = lam_best, type = "response"))
  roc_best <- pROC::roc(res$y, p, quiet = TRUE)
  auc_best <- as.numeric(pROC::auc(roc_best))
  
  nz <- nz[order(abs(nz[, 1]), decreasing = TRUE), , drop = FALSE]
  
  res$lambda_sel <- lam_best
  res$coef_best  <- nz
  res$auc_best   <- auc_best
  res$roc_best   <- roc_best
  res
}

## ================================================================
## SIMULATION: VARY SEEDS 400–500 AND RECORD SELECTED PCs
## ================================================================
if (!exists("merged_data")) {
  stop("merged_data must exist in the environment.")
}

seeds <- 1:500   # 100 runs

sel_vul_list <- vector("list", length(seeds))
sel_haz_list <- vector("list", length(seeds))
sel_cap_list <- vector("list", length(seeds))
names(sel_vul_list) <- names(sel_haz_list) <- names(sel_cap_list) <- as.character(seeds)

for (i in seq_along(seeds)) {
  s <- seeds[i]
  cat("Running seed =", s, "\n")
  
  ## ----------------- Train/test split (seed s) -----------------
  set.seed(s)
  train_idx <- caret::createDataPartition(merged_data$coliform, p = 0.6, list = FALSE)
  train_dat <- merged_data[train_idx, , drop = FALSE]
  test_dat  <- merged_data[-train_idx, , drop = FALSE]  # not used for selection
  
  ## ----------------- Module variable lists ---------------------
  if (exists("vul_tab")) {
    vars_vul <- setdiff(names(vul_tab), "grid_id")
  } else {
    vars_vul <- grep("^VUL_", names(train_dat), value = TRUE)
  }
  if (exists("haz_tab")) {
    vars_haz <- setdiff(names(haz_tab), "grid_id")
  } else {
    vars_haz <- grep("^HAZ_", names(train_dat), value = TRUE)
  }
  if (exists("cap_tab")) {
    vars_cap <- setdiff(names(cap_tab), "grid_id")
  } else {
    vars_cap <- grep("^CAP_", names(train_dat), value = TRUE)
  }
  
  ## ----------------- Penalty scheme for Vulnerability ----------
  vul_names <- intersect(vars_vul, names(train_dat))
  pc_id_raw <- suppressWarnings(as.integer(sub(".*PC(\\d+).*", "\\1", vul_names)))
  pf_vul <- if (all(!is.na(pc_id_raw)) && length(pc_id_raw) > 0L) {
    lo <- min(pc_id_raw); hi <- max(pc_id_raw)
    1 + (pc_id_raw - lo) / (max(1e-8, hi - lo))
  } else if (length(vul_names) > 0L) {
    rep(1, length(vul_names))
  } else {
    numeric(0)
  }
  if (length(pf_vul) > 0L) names(pf_vul) <- vul_names
  
  ## ----------------- Fit per-module LASSO on train_dat ---------
  # Vulnerability: β >= 0
  res_vul <- if (length(vars_vul) > 0L) {
    fit_lasso_module_pc_only(
      df = train_dat,
      module_vars = vars_vul,
      module_name = "Vulnerability",
      lower.limits = 0, upper.limits = Inf,
      dfmax = 5, pmax = 29,
      penalty.factor = pf_vul,
      use_1se = TRUE, target_k = 5
    )
  } else NULL
  res_vul <- pick_lambda_per_module(res_vul, lower.limits = 0, upper.limits = Inf)
  
  # Hazard: β >= 0
  res_haz <- if (length(vars_haz) > 0L) {
    fit_lasso_module_pc_only(
      df = train_dat,
      module_vars = vars_haz,
      module_name = "Hazard",
      lower.limits = 0, upper.limits = Inf,
      dfmax = 5, pmax = 6,
      use_1se = TRUE, target_k = 5
    )
  } else NULL
  res_haz <- pick_lambda_per_module(res_haz, lower.limits = 0, upper.limits = Inf)
  
  # Capacity: β <= 0
  res_cap <- if (length(vars_cap) > 0L) {
    fit_lasso_module_pc_only(
      df = train_dat,
      module_vars = vars_cap,
      module_name = "Capacity",
      lower.limits = -Inf, upper.limits = 0,
      dfmax = 5, pmax = 10,
      use_1se = TRUE, target_k = 5
    )
  } else NULL
  res_cap <- pick_lambda_per_module(res_cap, lower.limits = -Inf, upper.limits = 0)
  
  ## ----------------- Record selected PCs -----------------------
  sel_vul_list[[i]] <- if (!is.null(res_vul) && !is.null(res_vul$coef_best)) {
    rownames(res_vul$coef_best)
  } else character(0)
  
  sel_haz_list[[i]] <- if (!is.null(res_haz) && !is.null(res_haz$coef_best)) {
    rownames(res_haz$coef_best)
  } else character(0)
  
  sel_cap_list[[i]] <- if (!is.null(res_cap) && !is.null(res_cap$coef_best)) {
    rownames(res_cap$coef_best)
  } else character(0)
}

## ================================================================
## MAJORITY VOTING TABLES
## ================================================================
make_vote_table <- function(sel_list, seeds) {
  pcs <- unlist(sel_list, use.names = FALSE)
  if (length(pcs) == 0L) {
    return(data.frame(PC = character(0), count = integer(0), prop = numeric(0)))
  }
  tab <- sort(table(pcs), decreasing = TRUE)
  data.frame(
    PC    = names(tab),
    count = as.integer(tab),
    prop  = as.integer(tab) / length(seeds),
    row.names = NULL
  )
}

vote_vul <- make_vote_table(sel_vul_list, seeds)
vote_haz <- make_vote_table(sel_haz_list, seeds)
vote_cap <- make_vote_table(sel_cap_list, seeds)

maj_thresh <- length(seeds) / 2

maj_vul <- subset(vote_vul, count > maj_thresh)
maj_haz <- subset(vote_haz, count > maj_thresh)
maj_cap <- subset(vote_cap, count > maj_thresh)

cat("\n=== Majority PCs (selected in > 50% of runs) ===\n")

cat("\n[Vulnerability]\n")
print(maj_vul)

cat("\n[Hazard]\n")
print(maj_haz)

cat("\n[Capacity]\n")
print(maj_cap)

cat("\nTop PCs by frequency (no threshold) — Vulnerability:\n")
print(head(vote_vul, 5))

cat("\nTop PCs by frequency (no threshold) — Hazard:\n")
print(head(vote_haz, 5))

cat("\nTop PCs by frequency (no threshold) — Capacity:\n")
print(head(vote_cap, 5))

## ================================================================
## 2. Helpers
## ================================================================

## ---- Case-insensitive PC name matcher: "vul_PC1" / "VUL_PC1" both OK ----
get_pc_vars <- function(prefix, pc_index_vec, df) {
  all_names <- names(df)
  target_names <- paste0(prefix, "_PC", pc_index_vec)
  # match ignoring case, return the actual column names from df
  match_idx <- which(tolower(all_names) %in% tolower(target_names))
  all_names[match_idx]
}

## ---- Simple numeric median imputation ----
impute_numeric_median <- function(df) {
  out <- df
  for (v in names(out)) {
    if (is.numeric(out[[v]])) {
      med <- suppressWarnings(median(out[[v]], na.rm = TRUE))
      if (!is.finite(med)) med <- 0
      out[[v]][is.na(out[[v]])] <- med
    }
  }
  out
}

## ---- Constrained logistic regression with λ ≈ 0 (no effective penalty) ----
fit_constrained_glmnet_nopenalty <- function(
    df_train, df_test, y_name,
    pc_vars,
    lower_lim, upper_lim,
    lambda_small = 1e-10,
    module_label = "module"
) {
  if (length(pc_vars) == 0L) {
    stop(sprintf("No requested PCs found in training data for %s.", module_label))
  }
  
  # Keep only outcome + PCs
  df_tr <- df_train[, c(y_name, pc_vars), drop = FALSE]
  df_te <- df_test[,  c(y_name, pc_vars), drop = FALSE]
  
  # Impute numeric vars
  df_tr <- impute_numeric_median(df_tr)
  df_te <- impute_numeric_median(df_te)
  
  y_tr <- df_tr[[y_name]]
  y_te <- df_te[[y_name]]
  
  X_tr <- as.matrix(df_tr[, pc_vars, drop = FALSE])
  X_te <- as.matrix(df_te[, pc_vars, drop = FALSE])
  
  # Fit logistic regression with tiny λ and sign constraints
  fit <- glmnet(
    x = X_tr, y = y_tr,
    family = "binomial",
    alpha = 0,                 # ridge structure, but λ is tiny
    lambda = lambda_small,     # approximate "no penalty"
    standardize = TRUE,
    lower.limits = lower_lim,
    upper.limits = upper_lim
  )
  
  # Predictions and AUC
  p_tr <- as.numeric(predict(fit, newx = X_tr, type = "response"))
  p_te <- as.numeric(predict(fit, newx = X_te, type = "response"))
  
  auc_tr <- as.numeric(pROC::auc(pROC::roc(y_tr, p_tr, quiet = TRUE)))
  auc_te <- as.numeric(pROC::auc(pROC::roc(y_te, p_te, quiet = TRUE)))
  
  coefs <- as.matrix(coef(fit))
  
  cat("\n=====================================\n")
  cat(module_label, " — constrained logistic regression (λ ≈ 0)\n")
  cat("=====================================\n")
  cat("Train AUC:", round(auc_tr, 3),
      " | Test AUC:", round(auc_te, 3), "\n\n")
  cat("Coefficients:\n")
  print(round(coefs, 6))
  
  invisible(list(
    fit       = fit,
    coef      = coefs,
    auc_train = auc_tr,
    auc_test  = auc_te
  ))
}

## ================================================================
## 3. Vulnerability module (β ≥ 0)
##    PCs: 1, 9, 10, 15, 28
## ================================================================
vul_pc_idx <- c(1, 10)
vul_pcs <- get_pc_vars("vul", vul_pc_idx, train_dat)  # matches vul_PC*, VUL_PC*, etc.

fit_vul <- fit_constrained_glmnet_nopenalty(
  df_train     = train_dat,
  df_test      = test_dat,
  y_name       = "coliform",
  pc_vars      = vul_pcs,
  lower_lim    = 0,      # β ≥ 0
  upper_lim    = Inf,
  module_label = "Vulnerability (β ≥ 0; PCs 1, 9, 10, 15, 28)"
)

## ================================================================
## 4. Hazard module (β ≥ 0)
##    PCs: 1, 2, 3, 5, 6
## ================================================================
haz_pc_idx <- c(5, 6)
haz_pcs <- get_pc_vars("haz", haz_pc_idx, train_dat)

fit_haz <- fit_constrained_glmnet_nopenalty(
  df_train     = train_dat,
  df_test      = test_dat,
  y_name       = "coliform",
  pc_vars      = haz_pcs,
  lower_lim    = 0,      # β ≥ 0
  upper_lim    = Inf,
  module_label = "Hazard (β ≥ 0; PCs 1, 2, 3, 5, 6)"
)

## ================================================================
## 5. Capacity module (β ≤ 0)
##    PCs: 2, 4, 8, 9, 10
## ================================================================
cap_pc_idx <- c(2, 4, 8, 10)
cap_pcs <- get_pc_vars("cap", cap_pc_idx, train_dat)

fit_cap <- fit_constrained_glmnet_nopenalty(
  df_train     = train_dat,
  df_test      = test_dat,
  y_name       = "coliform",
  pc_vars      = cap_pcs,
  lower_lim    = -Inf,   # β ≤ 0
  upper_lim    = 0,
  module_label = "Capacity (β ≤ 0; PCs 2, 4, 8, 9, 10)"
)
```

## Two-sample tests

Next, we compare the performance of two-sample tests across three approaches for constructing the composite risk score: (1) supervised using PCs, (2) supervised using the original non-PC variables, and (3) unsupervised dimension reduction. We conduct one-sided \\textit{t}-tests, permutation \\textit{t}-tests, and Wilcoxon rank-sum tests to evaluate whether the mean or median risk score in the detect group is greater than that in the non-detect group, using a significance level of 0.05. A rainbow plot is also provided to visualize the distribution of the risk scores.

### Supervised (PCs)

In the following results, we observe significant differences in the composite risk scores between the detect and non-detect groups.

```{r}
# 1) Drop geometry and standardize column names
vul_tab <- sf::st_drop_geometry(vul_dat) |> dplyr::rename(index_vulnerability = index)
haz_tab <- sf::st_drop_geometry(haz_dat) |> dplyr::rename(index_hazard        = index)
cap_tab <- sf::st_drop_geometry(cap_dat) |> dplyr::rename(index_capacity      = index)

# 2) Build a compact risk table (align by grid_id), then min–max scale to [0,1]
risk_df <- vul_tab |>
  dplyr::select(grid_id, index_vulnerability) |>
  dplyr::left_join(haz_tab |> dplyr::select(grid_id, index_hazard),   by = "grid_id") |>
  dplyr::left_join(cap_tab |> dplyr::select(grid_id, index_capacity), by = "grid_id") |>
  dplyr::mutate(
    risk_raw = index_vulnerability + index_hazard - index_capacity,
    # guard against non-finite and all-NA edge cases
    risk_raw = dplyr::if_else(is.finite(risk_raw), risk_raw, NA_real_),
    risk_scaled = (risk_raw - min(risk_raw, na.rm = TRUE)) /
      (max(risk_raw, na.rm = TRUE) - min(risk_raw, na.rm = TRUE))
  ) |>
  dplyr::select(grid_id, risk_raw, risk_scaled)

# 3) Now join risk into your merged_data (or join modules later as you wish)
merged_data <- data |>
  dplyr::left_join(risk_df, by = "grid_id")

merged_data <- merged_data %>%
  # remove duplicate join suffixes
  dplyr::select(
    -tidyselect::matches("\\.x$"),
    -tidyselect::matches("\\.y$"),
    -tidyselect::matches("\\.x\\.x$"),
    -tidyselect::matches("\\.y\\.y$")
  ) %>%
  # drop obvious non-predictors / outcomes
  dplyr::select(
    -tidyselect::any_of(c(
      "System_Name", "Street_Address", "City", "State", "Zip",
      "Street_Address_Concat", "EIN", "StarLiMS_ID",
      "Date_Collected", "Sampling_Point", "Comment",
      "Accuracy", "Accuracy_Type", "grid_id",
      "risk", "Total_Coliform_Result", "Ecoli_Result"
    ))
  ) %>%
  # drop geometry if still present
  sf::st_drop_geometry() %>%
  # drop all A_* variables and specific unwanted columns
  dplyr::select(
    -tidyselect::starts_with("A_"),
    -tidyselect::any_of(c(
      "Ecoli_Result_Binary",
      "Coliform_Result_Binary",
      "ecoli",
      "y_fake"
    ))
  )

## ===============================================================
## Compute risk and compare by coliform status (one-sided: greater)
## ===============================================================

# Compute the risk score on merged data
risk <- merged_data$risk_scaled

# Split into training and test sets
risk.test <- risk[-train_idx]

# Basic parametric t-test (for reference)
t.test(risk.test ~ test_dat$coliform, alternative = "less")


library(ggplot2)
library(ggdist)
library(dplyr)

## Build plotting data (align, drop NAs)
df_plot <- tibble(
  risk = risk.test,
  coliform = test_dat$coliform
) %>%
  filter(!is.na(risk), !is.na(coliform)) %>%
  mutate(
    coliform = factor(coliform, levels = c(0, 1),
                      labels = c("Negative", "Positive"))
  )

## Half-eye + boxplot + jitter, grouped by coliform
ggplot(df_plot, aes(x = coliform, y = risk, color = coliform)) +
  ggdist::stat_halfeye(
    aes(fill = coliform),
    adjust = 0.5, width = 0.6, .width = 0, justification = -0.3
  ) +
  geom_boxplot(width = 0.25, outlier.shape = NA) +
  geom_point(size = 1.5, alpha = 0.2,
             position = position_jitter(seed = 1, width = 0.1)) +
  scale_x_discrete(NULL) +
  labs(y = "Risk score", color = "Coliform", fill = "Coliform") +
  theme_minimal() +
  theme(legend.position = "none")


## ===============================================================
## Nonparametric comparisons (KS, Wilcoxon, Permutation t-test)
## ===============================================================

# Extract vectors (drop NAs)
risk_vec <- risk.test
group_vec <- test_dat$coliform
keep <- !is.na(risk_vec) & !is.na(group_vec)
risk_vec <- risk_vec[keep]
group_vec <- as.factor(group_vec[keep])

# Split into two groups: 0 = negative, 1 = positive
risk_0 <- risk_vec[group_vec == 0]
risk_1 <- risk_vec[group_vec == 1]

## ---------------------------------------------------------------
## 1. Wilcoxon rank-sum test (one-sided: risk_1 > risk_0)
## ---------------------------------------------------------------
wilcox_res <- wilcox.test(risk_vec ~ group_vec,
                          alternative = "less",
                          exact = FALSE)
print(wilcox_res)

## ---------------------------------------------------------------
## 2. Permutation t-test (one-sided: mean(risk_1) > mean(risk_0))
## ---------------------------------------------------------------
set.seed(2025)
n_perm <- 10000  # number of permutations
obs_diff <- mean(risk_1) - mean(risk_0)

perm_diffs <- replicate(n_perm, {
  shuffled <- sample(group_vec)
  mean(risk_vec[shuffled == 1]) - mean(risk_vec[shuffled == 0])
})

# one-sided p-value: proportion of permuted diffs ≥ observed
p_perm <- mean(perm_diffs >= obs_diff)

cat("\nPermutation t-test (alternative = 'greater'):\n",
    "Observed mean difference =", round(obs_diff, 4),
    "| p-value =", signif(p_perm, 4), "\n")
```
